2.2.a Preprocessing and Monthly Positivity Figures

### 2024-03-18 2_scripts/2.1.1 Canpath Data.qmd
Update fig code for Canpath

table(ser_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
table(admin_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
```
#### Data quality check of those small dataset
```{r}
# admin_s: missing
table(is.na(admin_s$rid), useNA = 'ifany')     # rid no missing, not all unique (3 duplicates)
# remove duplicates: drop the earliest sample
admin_s<-admin_s[!admin_s$sid %in% c("DB0006218", "DB0009025", "DB0009310"),]

@@

# when those assay was collected?
m2 %>%
  group_by(region) %>%
  summarise(Freq=n(),
            tcol_min = min(date_col, na.rm=T),
            tcol_max = max(date_col, na.rm=T))
```

#### 1.3 save the data
```{r}
Canpaths<-m2
rm(m1, m2, n_admin, ser_sw, n_ser, qx_mtpmiss, ser_rid_dup)
rm(admin_s, qx_s, qx_s_mtp, ser_s)
save(Canpaths, file = "../1_data/private/CANPATH/Canpaths.RData")

# Save to the shared folder 
save(Canpaths, file = "~/../../jmy/CANPATH/Canpaths.RData")
```

>>>>

table(ser_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
table(admin_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
```

#### Data quality check of those small dataset
```{r}
# admin_s: missing
table(is.na(admin_s$rid), useNA = 'ifany')     # rid no missing, not all unique (3 duplicates)
# remove duplicates: drop the earliest sample
admin_s<-admin_s[!admin_s$sid %in% c("DB0006218", "DB0009025", "DB0009310"),]

@@

# when those assay was collected?
m2 %>%
  group_by(region) %>%
  summarise(Freq=n(),
            tcol_min = min(date_col, na.rm=T),
            tcol_max = max(date_col, na.rm=T))
```
#### 1.3 Add additional variables
```{r}
# add race variable: the coding on race is a bit messy in Canpath
# 14 variables were used to indicate diff racial minorities
# The cleaning code is provided by M.Knight
m2<-merge(m2, select(Canpath_qx, "ResearcherID",
          "C1_SDC_EB_ARAB","C1_SDC_EB_BLACK","C1_SDC_EB_CHINESE",
         "C1_SDC_EB_FILIPINO","C1_SDC_EB_JAPANESE",
         "C1_SDC_EB_KOREAN","C1_SDC_EB_LATIN","C1_SDC_EB_S_ASIAN",
         "C1_SDC_EB_SE_ASIAN","C1_SDC_EB_W_ASIAN","C1_SDC_EB_WHITE",
         "C1_SDC_EB_OTHER", "C1_SDC_EB_OTHER_OTSP","C1_SDC_EB_CA"),
         by.x = 'rid', by.y ="ResearcherID", 
         all.x = T, all.y = F

)

# Reorder the variables to fit Matthew's code:
m3<-cbind.data.frame(m2[,1:8], m2[,17:30], m2[,9:16])
m2<-m3
rm(m3)

# If possible, add a short para in the discussion section explaining why we can not categorize the indigenous people. Consult Yuan's email for more detail.
```

```{r}
# define whites and minorities:

#Racialized minorities - no indigenous responses included (applied later in for loop)
rm_indn<-c("Afro Latino Caribbean decent","East Indian descent, born in Kenya","Indo-Caribbean","white/asian",
           "SE Asian (Taiwanese)","i am South African, I carry in my veins European, African, Asian and indigenous Southern Afircan",
           "West Indian from Guyana","Guyanese - West Indian","South American","caucasian - european, indian, iraqi",
           "Ugandian of South Asian heritage","Trinidadian","White, Chinese, West Asian", "Japanese/Irish","Chinese, Indian, English",
           "East Indian born in Guyana","West-Indian","South African of mixed races","half Japanese half White",
           "Armenian","South Asian from East Africa","Indo Caribbean","Japanese and European","Black/white",
           "Mixed race Chinese-Guyanese-White","East Asian","west indian",
           "White, Indigeneous, Latin-American","Indian born in the Caribbean.",
           "White and Khoisan","European Hispanic","Armenian mother, Irish father, but born in England",
           "Persian","Guyanese,south asian descent","Mixed heritage (white and black)","Caribbean, Mixed Race",
           "Assyrian","Caucasian African","WEST INDIAN/CHINESE/","Latino and European","Of Indian decent - Born in Caribbean",
           "White/Southeast Asian","mixed european/chinese","Indo-caribbean","Mixed Race (Black, White, Hispanic,Soth Asian)",
           "South Asian/ Mixed South african","Canadian, German & Caribbean","Black Canadian",
           "East Indian","Mixed European & Southeast Asian","Je suis Afro Latino-AmÃ\u0083Â\u0083Ã\u0082Â©ricain","latino-amÃ\u0083Â\u0083Ã\u0082Â©ricain occidental",
           "Nord africain","Berbere","Berbere (Afrique du nord)","Berbere Nord Africain","amazigh","maghrebine","Nord-Africain",
           "Juif marocain","Nord africaine","90% white 10%Jewish-Arab","Canadian and Afrikaans",
           "mixed with Jamaican", "guianese of east indian descent",
           "caribbean chinese", "chinese and french background and other",
           "arabe-africaine", "berbère","persian",
           "indo-european","colombienne","japanese & swedish",
           "east indian","middle east",
           "armenian",
           "north african","south american, african, chinese","middle eastern",
           "middle east, and central europe","cherokee",
           "irish and guyanese","african","zimbawayan",
           "dutch indonesian",
           "trinidadian","south america - biracial",
           "punjabi","west indian of east indian descent",
           "mixed race - mother part black/father white caucasian","jamaican",
           "canadian chinese","cambodgienne","africain","asian","mauricien",
           "taiwanese","middle east/ persion",
           "1/4 middle eastern (syria which is now lebanon) from paternal grandfather",
           "mixed afghan and english","tunisien",
           "jamaican.","west  indian","zoroastrian (ancient persian ancestry)",
           "white/west asian","kabyle (berbère) d'afrique du nord","white mexican jewish")

#Racialized minorities - indigenous responses included
rm_indy<-c("Metis/European descent","White and aboriginal","Metis","Indigenous / European","Metis French Canadian",
           "Metis Canadian","Blanche ascendance micmak","Metis ancestry","white a bit of indigenous","métis",
           "french metis", "30% indigenous metis/70% white","metis",
           "culturally white, 3/16 aboriginal","mik'maw","inuit / european", "autochtone",
           rm_indn)

#White or Caucasian
w<-c("white newfoundlander","White British",
     "White passing","Caucasian","White - British Decent",
     "White, British descent","Caucasian white", "White - Canadian Irish descent",
     "white french canadian", "Caucasian female","White - Canadian","caucasian","White Canadian Scottish parents","Canadian Caucasian","White British Descent",
     "Canadian - White - 4 generations",
     "white, British descent","CAUCASIAN",
     "Caucasian - European descent","Canadian white Caucasian","white Canadian" ,
     "Canadian White","White Franco Ontarian","Caucasian Canadian",
     "white  Canadian       North American", "White Canadian born","White Canadian, Parents of Ukrainian and French background",
     "White Anglo-Saxon","White/Cdn born/European grandparents",
     "Canadian, white","White caucasian","white not Europe.",
     "Hebrew (Jewish) Canadian white","White (Canadian)","White","White from Canada",
     "White  English decent=","White bilingual Canadian - Family from Quebec","white canadian","White  Canadian",
     "White Canadian, no known ethnicities","White (Causcasian)","white","blanc Canadienne","Canadienne blanche",
     "BLANCHE","Blanc.", "Blanc (Canada)","Canadienne blanc","Caucasien","Blanc - Canadien",
     "Blanc canadienne","Blanc du canada","Blanc canadien","Blanc canadian",
     "Blanche (Canadienne)","une femme de race blanche","blanc","Blanche","Blanc canada","Blanc","Blanche, Canadienne",
     "blanc quebecois","caucasien",
     "White & Canadian","White Canadian","blanc du canada","White- true blue CANADIAN","White Scottish Descent",
     "blanche canadienne","Blanc Canadien","canadien francais blanc",
     "white, scottish and other",
     "ccaucasian","caucasienne",
     "white european","white, mainly scottish/english/irish",
     "white, of european decent with mi'kmaq ancestors several generations removed.",
     "canadian caucasian", "white, norwegian and scottish",
     "canadien francais, caucasien","european caucasian",
     "white celtic")
```

```{r}
# a for loop for creating race var:
m2$race<-NULL

for(i in 1:nrow(m2)){
  m2$race[i]<-case_when(
    #select pnts and nothing else
    !is.na(m2[i,22]) & 
      sum(is.na(m2[i,9:22])) == 13 ~ "pnts",

    #select only 1 white ethnicity
    !is.na(m2[i,19]) & 
      sum(is.na(m2[i,9:22])) == 13 ~ "White",

    #select only 1 racialized minority ethnicity
    sum(!is.na(m2[i,9:18])) == 1 &
      sum(is.na(m2[i,9:22])) == 13 ~ "Racialized minority",

    #select 1 white ethnicity and at least 1 racialized minority ethnicity
    sum(!is.na(m2[i,9:18])) > 0 & 
      !is.na(m2[i,19]) &
      sum(is.na(m2[i,22])) ~ "Racialized minority",

    #select 2+ racialized minority ethnicities
    sum(!is.na(m2[i, 9:18])) > 1 &
      sum(is.na(m2[i, c(19, 22)])) == 2 ~ "Racialized minority",

    #select at least 1 racialized minority ethnicity and at least 1 "other" ethnicity
    sum(!is.na(m2[i, 9:18])) > 0 &
      sum(is.na(m2[i, c(19, 22)])) == 2 &
      sum(!is.na(m2[i, 20,21])) > 0 ~ "Racialized minority",

    #select white, but also identify as a racialized minority using text box
    sum(is.na(m2[i, c(9:18, 22)])) == 11 &
      !is.na(m2[i,19]) &
      sum(!is.na(m2[i, c(20:21)])) >= 1 &
      m2[i, c(21)] %in% (rm_indn) ~ "Racialized minority",

    #select white, but also identify as white using text box
    sum(is.na(m2[i, c(9:18, 22)])) == 11 &
      !is.na(m2[i,19]) &
      sum(!is.na(m2[i, c(20:21)])) >= 1 &
      m2[i, c(21)] %in% (w) ~ "White",

    #select other and identify as a racialized minority using text box
    sum(is.na(m2[i, c(9:19, 22)])) == 12 &
      m2[i, 21] %in% (rm_indn) ~ "Racialized minority",

    #select other and identify as white using text box
    sum(is.na(m2[i, c(9:19, 22)])) == 12 &
      m2[i, 21] %in% (w) ~ "White",

    TRUE ~ NA)
}

table(m2$race, useNA = 'ifany')

# pnt = prefer not to answer
table(m2[,22], m2$race, useNA = 'ifany')

# remove loose race vars
m2<-select(m2, -c(9:22))
```

#### 1.4 save the data
```{r}
Canpaths<-m2
rm(m1, m2, n_admin, ser_sw, n_ser, qx_mtpmiss, ser_rid_dup)
rm(admin_s, qx_s, qx_s_mtp, ser_s)
save(Canpaths, file = "../1_data/private/CANPATH/Canpaths.RData")

# Save to the shared folder 
save(Canpaths, file = "~/../../jmy/CANPATH/Canpaths.RData")
# Save to the shared folder as .csv format
write.csv(Canpaths, file = "~/../../jmy/CANPATH/Canpaths.csv", row.names=FALSE)
```

### 2024-03-18 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
Update fig code for Canpath

@@

p01<-CBS_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=Adj, colour=Adj)) +
  geom_point(aes(colour=Adj)) +
  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
#  geom_line(aes(colour=Adj))  + #non-smooth curve
  theme_bw()

@@

>>>>

**Notes:** In Canpath data we don't have to worry about weighting, as one people has been sampled only once, then tested by multiple asays (either Gingras x3 or Abbott x2)

##### 2.3.1 Pre-process Canpath data
```{r}
Canpaths %>%
  select(region, assay_id) %>%
  group_by(region, assay_id) %>%
  summarise(Freq=n())

# remove 9 AB cases with Gingras abti-N (Mount Sinai Hosp)
Canpaths<-Canpaths %>%
  .[!(.$region=="Alberta" & .$assay_id=="10010-00"),]

# remove 2 BC cases with Abbott anti-N
Canpaths<-Canpaths %>%
  .[!(.$region=="BC" & .$assay_id=="20010-03"),]

# create samplemonth var
Canpaths$samplemonth<-as.yearmon(Canpaths$date_col)
table(Canpaths$samplemonth, useNA = 'ifany')

# see samplemonth by assay
Canpaths %>%
  .[!is.na(.$region),] %>%
  select(samplemonth, assay_id) %>%
  group_by(samplemonth, assay_id) %>%
  summarise(Freq=n())
```

```{r}
# create sero-prevalance tab
tb1<-Canpaths%>%
  group_by(samplemonth, region, .add=TRUE) %>%
  summarise(n = n(), )

tb2<-Canpaths[Canpaths$pos==1,] %>%
  group_by(samplemonth, region, .add=TRUE) %>%
  summarise(n.pos = n(), )

tb3<-merge(tb1, tb2, by = c("samplemonth", "region"), all.x = T, all.y = F)
tb3$prop.n<-tb3$n.pos/tb3$n
tb3$prop.n[is.na(tb3$prop.n)]<-0

Canpaths_pr<-tb3[!is.na(tb3$region) &
                 !is.na(tb3$samplemonth),]

# Third-party sens & spec: Abbott-N
# Sen=92.7%, spec = 99.9% 
# Source: https://assets.publishing.service.gov.uk/media/5f05887ad3bf7f2be6e02161/Evaluation__of_sensitivity_and_specificity_of_4_commercially_available_SARS-CoV-2_antibody_immunoassays.pdf
Canpaths_pr$prop.n_RG<-rep(NA, 38)
Canpaths_pr[Canpaths_pr$region=="Alberta",]$prop.n_RG<- 
  (Canpaths_pr[Canpaths_pr$region=="Alberta",]$prop.n+0.999-1)/(0.999+0.927-1)

# Third-party sens & spec: Mount Sinai Hosp (Gingras), from T.Murphy
# Sen=83.5%, spec = 96.4%
Canpaths_pr[!Canpaths_pr$region=="Alberta",]$prop.n_RG<- 
  (Canpaths_pr[!Canpaths_pr$region=="Alberta",]$prop.n+0.999-1)/(0.999+0.927-1)

## According to Yuan's request, add sens & spec to all the obs
Canpaths_pr$sens[Canpaths_pr$region=="Alberta"]<-0.927
Canpaths_pr$sens[!Canpaths_pr$region=="Alberta"]<-0.835
Canpaths_pr$spec[Canpaths_pr$region=="Alberta"]<-0.999
Canpaths_pr$spec[!Canpaths_pr$region=="Alberta"]<-0.964

Canpaths_pr$data<-rep("Canpaths", length(Canpaths_pr$samplemonth))

@@

p01<-CBS_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=Adj, colour=Adj)) +
  geom_point(aes(colour=Adj)) +
#  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
  geom_line(aes(colour=Adj))  + #non-smooth curve
  theme_bw()

@@

#### 3.2 Canpath: make plots by region
```{r}
# Reshape meltted data
Canpaths_melt<-reshape2::melt(select(Canpaths_pr, samplemonth, region,
                                       prop.n, prop.n_RG), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "Adj",
                        value.name = "prop")

p02<-Canpaths_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=Adj, colour=Adj)) +
  geom_point(aes(colour=Adj)) +
#  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
  geom_line(aes(colour=Adj))  + 
  theme_bw()+
  scale_y_continuous(limits = c(0,1))

p02 + facet_wrap(~region)
